Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-05 latex/main.tex

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{recombinants_over_time.pdf}
    \caption{Recombination events across the pandemic. (A) Timing of the recombination events (black curve, right y-axis) relative to the global case numbers (\cite{Dong2020}; left y-axis). Cases are color-coded by their Scorpio labels, with the key showing those groupings with $\geq 1,000$ samples in the Viridian dataset. Recombinants more commonly occur (B) in weeks with high case numbers and (C) in weeks with extensive lineage variation, as estimated by high expected heterozygosity; solid lines show fits by linear regression, allowing for non-zero intercepts to account for unknown cases and heterozygosity not captured by the Scorpio labels.}
    \label{fig:recombinants_over_time}
\end{figure}

\subsection*{Efficient inference and analysis of a pandemic-scale ARG}
The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
which overwhelmed existing computational approaches,
motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
the functionality required largely overlaps.
Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
and the VCF Zarr data storage format \cite{Czech2025}
to describe relationships among virus samples, building an ARG over time as data becomes available.
By reusing high-quality and high-performance components of these toolkits,
we could develop the sc2ts inference method itself and
perform all the pandemic-scale analyses in this article with Python code alone and
no compiled languages.

@@

\section*{SUPPLEMENTAL INFORMATION INDEX}

%%%  Supplemental information must be uploaded as
%%%  separate files. In the main text, please list the
%%%  files to be included in a brief index. For details,
%%%  please review the supplemental information guidelines:

%%%  Journals with STAR Methods:
%%%  https://www.cell.com/STAR-supplemental-information

\begin{description}
  \item Figures S1-SX and their legends in a PDF
  \item Table S1. A descriptive title for an Excel file that was too large to appear in the PDF
\end{description}

>>>>

\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{recombinants_over_time.pdf}
    \caption{Recombination events across the pandemic. (A) Timing of the
recombination events (black curve, right y-axis) relative to the global case
numbers (\cite{Dong2020}; left y-axis). Cases are color-coded by their Scorpio
labels, with the key showing those groupings with $\geq 1,000$ samples in the
Viridian dataset. Recombinants more commonly occur (B) in weeks with high case
numbers and (C) in weeks with extensive lineage variation, as estimated by high
expected heterozygosity; solid lines show fits by linear regression, allowing
for non-zero intercepts to account for unknown cases and heterozygosity not
captured by the Scorpio labels.}
    \label{fig:recombinants_over_time}
\end{figure}

\subsection*{Efficient inference and analysis of a pandemic-scale ARG}
The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
which overwhelmed existing computational approaches,
motivating the rapid development of new methods\cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
the functionality required largely overlaps.
Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
and the VCF Zarr data storage format \cite{Czech2025}
to describe relationships among virus samples, building an ARG over time as data becomes available.
By reusing high-quality and high-performance components of these toolkits,
we could develop the sc2ts inference method itself and
perform all the pandemic-scale analyses in this article with Python code alone and
no compiled languages.

@@

% \section*{SUPPLEMENTAL INFORMATION INDEX}

% %%%  Supplemental information must be uploaded as
% %%%  separate files. In the main text, please list the
% %%%  files to be included in a brief index. For details,
% %%%  please review the supplemental information guidelines:

% %%%  Journals with STAR Methods:
% %%%  https://www.cell.com/STAR-supplemental-information

% \begin{description}
%   \item Figures S1-SX and their legends in a PDF
%   \item Table S1. A descriptive title for an Excel file that was too large to appear in the PDF
% \end{description}
